Measuring contrast processing in the visual system using SSVEP

### 2024-04-07 review.qmd
Started figure summarising clinical differences

First, we implement a simple non-linear transducer with a single input to illustrate the typical sigmoidal shape of a contrast response function [@fig-contrastresponsefunction].

Next, we show that adding a second component which contributes to the gain pool of the first, results in a complex pattern of intermodulation terms [@fig-plotspectra].

@@

To illustrate the effect of contrast gain control [@Heeger1992], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT [@fig-intermodulation].

@@

Autism is a condition often associated with differences in vision [@Simmons2009] and other senses [@MacLennan2022]. @Pei2014 used a sweep-VEP method with counterphase flickering stimuli, and found weaker responses in autistic children at spatial frequencies around 8c/deg, compared with age-matched controls. This was subsequently replicated in a further pediatric sample by @Vilidaite2018, who additionally found weaker responses in autistic adults at the second harmonic (using on/off flicker). Interestingly this study replicated its key findings in a *drosophila* genetic model of autism (*Nhe3* mutations), illustrating the translational potential of the steady-state approach, as well as identifying a possible biomarker for autism.

@@

(maybe here we want a 4-panel figure plotting cartoon models of the four conditions - it would make for a good comparison, and we could potentially base it on data from the original studies)

>>>>

First, we implement a simple non-linear transducer with a single input to illustrate the typical sigmoidal shape of a contrast response function (@fig-contrastresponsefunction).

Next, we show that adding a second component which contributes to the gain pool of the first, results in a complex pattern of intermodulation terms (@fig-plotspectra).

@@

To illustrate the effect of contrast gain control [@Heeger1992], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT (@fig-intermodulation).

@@

Autism is a condition often associated with differences in vision [@Simmons2009] and other senses [@MacLennan2022]. @Pei2014 used a sweep-VEP method with counterphase flickering stimuli, and found weaker responses in autistic children at spatial frequencies around 8c/deg, compared with age-matched controls. This was subsequently replicated in a further pediatric sample by @Vilidaite2018 (see [@fig-plotclinical]c), who additionally found weaker responses in autistic adults at the second harmonic (using on/off flicker). Interestingly this study replicated its key findings in a *drosophila* genetic model of autism (*Nhe3* mutations), illustrating the translational potential of the steady-state approach, as well as identifying a possible biomarker for autism.

@@

```{r}

pdf('Figures/clinicalplot.pdf',width=12,height=10)
par(mfrow=c(2,2))

finecontrastsdB <- seq(-18,36,1)
finecontrasts <- 10^(finecontrastsdB/20)
ticklocsx <- seq(-18,36,6)
ticklabelsx <- c('1/8','1/4','1/2',1,2,4,8,16,32,64)

# fits to data from Tsai et al. (2011)
epilepsyparams <- matrix(0,nrow=2,ncol=4)
epilepsyparams[1,] <- c(3.4,0.15,1.4,1)  # controls
epilepsyparams[2,] <- c(5.3,0.23,1.4,1)  # patients

respEPIcontrols <- epilepsyparams[1,1] * ((finecontrasts/100)^epilepsyparams[1,3]/(epilepsyparams[1,2]^epilepsyparams[1,3] + (finecontrasts/100)^epilepsyparams[1,3])) + epilepsyparams[1,4]

respEPIpatients <- epilepsyparams[2,1] * ((finecontrasts/100)^epilepsyparams[2,3]/(epilepsyparams[2,2]^epilepsyparams[2,3] + (finecontrasts/100)^epilepsyparams[2,3])) + epilepsyparams[2,4]

ticklocsy <- seq(0,6,2)
ticklabelsy <- ticklocsy

par(pty="s")  # make axis square
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(-18,36), ylim=c(0,6))   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.3)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1)
title(xlab="Target contrast (%)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)   
title(ylab="SSVEP response", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(main='(a) Epilepsy',cex.main=2)

lines(finecontrastsdB,respEPIpatients,lwd=4,col='blue')
lines(finecontrastsdB,respEPIcontrols,lwd=4,lty=2,col='black')

legend(-18,6,c('Epilepsy patients','Control participants'),lwd=4,lty=c(1,2),col=c('blue','black'),box.lwd=2)

ticklocsy <- seq(0,6,2)
ticklabelsy <- ticklocsy

par(pty="s")  # make axis square
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(-18,36), ylim=c(0,6))   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.3)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1)
title(xlab="Target contrast (%)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)   
title(ylab="SSVEP response", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(main='(b) Amblyopia',cex.main=2)

# fits to data from Vilidaite et al (2018)
autismparams <- matrix(0,nrow=2,ncol=4)
autismparams[1,] <- c(0.6253,0.2701,125550,5.6199)  # ASD group
autismparams[2,] <- c(1.0131,0.3369,3589900,6.8622) # control group

respASD <- (autismparams[1,1]*(finecontrasts^autismparams[1,4])/(autismparams[1,3] + finecontrasts^autismparams[1,4])) + autismparams[1,2]
respCON <- (autismparams[2,1]*(finecontrasts^autismparams[2,4])/(autismparams[2,3] + finecontrasts^autismparams[2,4])) + autismparams[2,2]

ticklocsy <- seq(0,1.5,0.5)
ticklabelsy <- ticklocsy

par(pty="s")  # make axis square
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(-18,36), ylim=c(0,1.5))   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.3)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1)
title(xlab="Target contrast (%)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)   
title(ylab="SSVEP response", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(main='(c) Autism',cex.main=2)

lines(finecontrastsdB,respASD,lwd=4,col='darkgreen')
lines(finecontrastsdB,respCON,lwd=4,lty=2,col='black')

legend(-18,1.5,c('ASD children','Control children'),lwd=4,lty=c(1,2),col=c('darkgreen','black'),box.lwd=2)

ticklocsy <- seq(0,6,2)
ticklabelsy <- ticklocsy

par(pty="s")  # make axis square
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(-18,36), ylim=c(0,6))   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.3)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1)
title(xlab="Target contrast (%)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)   
title(ylab="SSVEP response", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(main="(d) Parkinson's",cex.main=2)

dev.off()

![caption.](Figures/clinicalplot.pdf){#fig-plotclinical}
